---
title: "Data Exploration"
format: html
---

## September 3, 2026 - work in progress - scratchpad script

```{r, echo=FALSE}
#| label: libraries


library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(lubridate)
#geographic packages
library(countrycode)
library(sf)
library(rnaturalearth)
library(shiny)
library(httr2)
library(jsonlite)
library(here)
library(renv)

```


```{r, echo=FALSE}
#| label: load_data

tm <- read_parquet("Data/processed/TM_staples.parquet")

```

#examine the data
```{r, echo=FALSE}
#: label: examine the data
#1) are there any countries with no data for these staples?
tm |>
  filter(Value==0) |>
  select(Year, Reporter.Countries, Element.Code, Item, Element) |>
  arrange(Year, Reporter.Countries)

#Looking at zeros by Element
tm |>
  group_by(Element) |>
  summarise(
    n = n(),
    n_zero = sum(Value == 0, na.rm = TRUE),
    pct_zero = round(100 * n_zero / n, 2),
    n_na = sum(is.na(Value))
  )

#flow consistency check from Claude
flows <- tm |>
  select(Reporter.Countries, Partner.Countries, Item, Year, Element, Value) |>
  pivot_wider(names_from = Element, values_from = Value)

flows |>
  summarise(
    export_qty0_val_pos = sum(`Export quantity` == 0 & `Export value` > 0, na.rm = TRUE),
    export_val0_qty_pos = sum(`Export value` == 0 & `Export quantity` > 0, na.rm = TRUE),
    import_qty0_val_pos = sum(`Import quantity` == 0 & `Import value` > 0, na.rm = TRUE),
    import_val0_qty_pos = sum(`Import value` == 0 & `Import quantity` > 0, na.rm = TRUE),
    total_rows = n()
  )


fq <- tm |>
   select(Reporter.Countries, Partner.Countries, Item, Year, Element, Value, Flag) |>
   pivot_wider(names_from = Element, values_from = c(Value, Flag))

cat("Export value (when quantity=0), quantiles:\n")
fq |> filter(`Value_Export quantity` == 0, `Value_Export value` > 0) |>
   pull(`Value_Export value`) |> quantile(probs = c(.5, .75, .9, .95, .99, 1)) |> print()
 
cat("\nExport quantity (when value=0), quantiles (tonnes):\n")
fq |> filter(`Value_Export value` == 0, `Value_Export quantity` > 0) |>
   pull(`Value_Export quantity`) |> quantile(probs = c(.5, .75, .9, .95, .99, 1)) |> print()
 
cat("\nCount of 'large' anomalies: Export qty>0 with value=0, but qty > 100 tonnes\n")
big_qty_zero_val <- fq |> filter(`Value_Export value` == 0, `Value_Export quantity` > 100)
 nrow(big_qty_zero_val)
 
cat("\nSample of large export-quantity/zero-value rows (potential food aid/data issue):\n")
big_qty_zero_val |>
   select(Reporter.Countries, Partner.Countries, Item, Year, `Value_Export quantity`) |>
   arrange(desc(`Value_Export quantity`)) |>
   slice_head(n = 10) |> print()


# Test the document's claim that "trade data are rounded to the nearest metric ton"
cat("Unit values used for quantity elements:\n")
tm |> filter(Element %in% c("Export quantity", "Import quantity")) |> distinct(Unit) |> print()

cat("\nSample of Export quantity values to check for sub-tonne decimals:\n")
tm |> filter(Element == "Export quantity", Value > 0, Value < 1) |>
  slice_head(n = 10) |>
  select(Reporter.Countries, Partner.Countries, Item, Year, Value)

```

### Some remaining data cleaning & data exploration questions for this exercise
1 Do we care about $1,000 or less of trade for a country for staples?
2 Do we care about 1 tonne or less of trade for a country for staples?
3 What about mis-matches? 
    a High quantity exports, low or no export value; 
    b Low quantity exports, high value
    c High quantity imports, low or no import value
    d Low quantity exports, low or no export value
4 How much of question 3 can be explained by questions 1-2?
5 Are there yearly averages for values per tonne of each staple and what is the distribution?


### To dos:
1 Need to download the food balance sheet to have the gross food supply as a percentage of reporter country data and use 1% as a meaningful cut-off (still trying to get a reference for this)
Several terms: gravity analysis, materiality filter, scarcity filter, etc.

2 Use the FBS to normalize and remove meaningless trade amounts